## Calculating scores from quantiles using the Johnson distribution

* [Notes](http://dushoff.github.io/notebook/qbd.html)
* [Code](https://github.com/dushoff/scratch/blob/v0.3/johnson.R)

To use, you can load the code.
```{r load}
source("https://raw.githubusercontent.com/dushoff/scratch/v0.3/johnson.R")
```

To evaluate a data point, you would use `jsqdens` to estimate the density at the observed point.

For example, if the prediction was 40 (50%CI, 7–122), and the observation was 195, you could construct a Johnson distribution on the log scale, and then project back and find the density of the distribution at the observed value, by saying:

```{r example}
print(jsqdens(obs=195, q=c(7, 40, 122), P=0.5, logq=TRUE))
```

$q$ is the prediction with CIs; $P=0.5$ represents 50% CIs, and `logq` is used to do the calculation on the log scale.

The log scale seems like the sensible one to use for incidece or peak size. We might use the untransformed scale for incidence; we might do something different for peak week, for example.

Since $q$ is passed as a separate parameter, the easiest way to implement this on a data frame would probably be to use `plyr`. I'm happy to keep working on this, if there is interest.
